---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$datetime <- with(activity, 
                          as.POSIXct(date, format = "%Y-%m-%d") 
                          + (interval * 60))
```


## What is mean total number of steps taken per day?
```{r}
stepsPerDay <- tapply(activity$steps[!is.na(activity$steps)],
                activity$date[!is.na(activity$steps)], sum)

hist(stepsPerDay, breaks = 10, xlab = "Total Steps Per Day",
     main = "Total number of steps taken each day")

mean(stepsPerDay)

median(stepsPerDay)
```


## What is the average daily activity pattern?
```{r}
avSteps <- tapply(activity$steps[!is.na(activity$steps)],
                  activity$interval[!is.na(activity$steps)],mean)

xAxis <- unique(activity$interval)

plot(xAxis, avSteps, main = "Daily Activity Pattern", 
     xlab = "5-minute intervals", ylab = "Average number of steps",
     type = "l")

mInterval <- as.numeric(names(which.max(avSteps)))
```
5 minute interval that contains the maximum number of steps on average across 
all days is 

```{r bestInterval, echo=FALSE}
mInterval
```


## Imputing missing values
Total number of of missing values in the dataset
```{r}
sum(!complete.cases(activity))
```

Since the values are missing not for random 5 minute intervals but for the whole 
days, let's populate the missing values with the rounded averages across all 
days for the correspoinding intervals

```{r}
activityFull <- activity
activityFull$steps[is.na(activity$steps)] <- 
        with(activity[is.na(activity$steps),], 
                round(avSteps[as.character(interval)])
             )
```

A histogram of the total number of steps taken each day for the fully populated 
dataset  

```{r}
stepsPerDayFull <- tapply(activityFull$steps, activity$date, sum) 

hist(stepsPerDayFull, breaks = 10, xlab = "Total Steps Per Day",
     main = "Total number of steps taken each day")

```  
Mean total number of steps taken per day.
```{r}
mean(stepsPerDayFull)
```
Median total number of steps taken per day.
```{r}
median(stepsPerDayFull)
```
## Are there differences in activity patterns between weekdays and weekends?
